---
title: Genome-wide analysis of heat shock response 
author: 
  - name: Peter Humburg
    affiliation: 1
address:
  - code: 1
    address: Wellcome Trust Centre for Human Genetics, University of Oxford, Roosevelt Dr., Oxford, OX3 7BN, UK
date: `r format(Sys.time(), "%a %d %b %Y")`
---

```{r, setup, include=FALSE}
library(gdata)
library(affy)
library(vsn)
library(limma)
library(scatterplot3d)
library(sparcl)
library(sva)

opts_chunk$set(autodep=TRUE)
opts_chunk$set(tidy=TRUE)
opts_chunk$set(echo=FALSE)
options(figcap.prefix = "Figure", figcap.sep = ":", figcap.prefix.highlight = "**")
options(tabcap.prefix = "Table", tabcap.sep = ":", tabcap.prefix.highlight = "**")
```


```{r, functions, include=FALSE}
pcaPlot <- function(data, col, pch, ...){
        pca <- prcomp(t(data))
        scores <- as.data.frame(pca$x)
        plot(scores$PC1, scores$PC2, col=col, pch=pch, ...)
}

figRef <- local({
    tag <- numeric()
    created <- logical()
    used <- logical()
    function(label, caption, prefix = options("figcap.prefix"), 
        sep = options("figcap.sep"), prefix.highlight = options("figcap.prefix.highlight")) {
        i <- which(names(tag) == label)
        if (length(i) == 0) {
            i <- length(tag) + 1
            tag <<- c(tag, i)
            names(tag)[length(tag)] <<- label
            used <<- c(used, FALSE)
            names(used)[length(used)] <<- label
            created <<- c(created, FALSE)
            names(created)[length(created)] <<- label
        }
        if (!missing(caption)) {
            created[label] <<- TRUE
            paste0(prefix.highlight, prefix, " ", i, sep, prefix.highlight, 
                " ", caption)
        } else {
            used[label] <<- TRUE
            paste(prefix, tag[label])
        }
    }
})

tabRef <- local({
    tag <- numeric()
    created <- logical()
    used <- logical()
    function(label, caption, prefix = options("tabcap.prefix"), 
        sep = options("tabcap.sep"), prefix.highlight = options("tabcap.prefix.highlight")) {
        i <- which(names(tag) == label)
        if (length(i) == 0) {
            i <- length(tag) + 1
            tag <<- c(tag, i)
            names(tag)[length(tag)] <<- label
            used <<- c(used, FALSE)
            names(used)[length(used)] <<- label
            created <<- c(created, FALSE)
            names(created)[length(created)] <<- label
        }
        if (!missing(caption)) {
            created[label] <<- TRUE
            paste0(prefix.highlight, prefix, " ", i, sep, prefix.highlight, 
                " ", caption)
        } else {
            used[label] <<- TRUE
            paste(prefix, tag[label])
        }
    }
})
``` 


# Introduction
The heat shock response is a highly conserved mechanism responsible for the 
maintenance of cellular function under stress. Originally identified as part of the 
response to increased environmental temperatures [@Ritossa62] heat shock proteins (HSPs) 
have since been implicated  in susceptibility to infection [@Kee08], cancer risk [@Sfar10] 
and drug response [@Martin04]. Despite its obvious significance many details of the
heat shock response still remain uncertain.

Here we study the genome-wide changes to gene expression induced by heat shock in
60 HapMap cell lines from Yoruban individuals [@Gibbs10] to identify genes and pathways
involved in the human heat shock response. To further elucidate underlying mechanisms
genetic variants modulating gene expression are mapped across all relevant genes.

# Material and Methods

## Quality Control

```{r, loadExpression, echo=FALSE, cache=TRUE}
## probe intensities
rawIntensity <- read.delim("TPS00590_FinalReport_12Oct09.txt", skip=8)

sig.col<-seq(3, 722, 6)
det.col<-seq(8, 722, 6)

sig.data.raw <- rawIntensity[,sig.col]
det.data.raw <- rawIntensity[,det.col]

row.names(sig.data.raw) <- rawIntensity$ProbeID
row.names(det.data.raw) <- rawIntensity$ProbeID

colnames(sig.data.raw) <- gsub(".AVG_Signal", "", colnames(sig.data.raw))
colnames(det.data.raw) <- gsub(".Detection.Pval", "", colnames(det.data.raw))

## sample information
n<-read.delim("YRI_sample list file_03NOV09.txt")

## remove empty lines at end of file
n<-n[1:120,] 
factors <- which(sapply(n, is.factor))
for(i in factors){
	n[[i]] <- factor(as.character(n[[i]]))
}


newNames <- paste(n[["Cell.Line"]], n$Treatment, sep="_")
names(newNames)<- n$ARRAY.CODE

## typo in sample ID
names(sig.data.raw)[13]<- "NM_75"
names(det.data.raw)[13]<- "NM_75"

names(sig.data.raw) <-gsub(".", "_", fixed=TRUE, names(sig.data.raw))
names(det.data.raw) <-gsub(".", "_", fixed=TRUE, names(det.data.raw)) #differing name format
names(sig.data.raw) <- newNames[names(sig.data.raw)]
names(det.data.raw) <- newNames[names(det.data.raw)]
```
Probe intensities for resting and stimulated cells were imported into R for further
processing. 

# Results
## Quality Control



# Appendix {-}
## Custom functions used
```{r, ref.label="functions", eval=FALSE, echo=TRUE}
```

## Session Info
```{r sessionInfo, echo=FALSE}
sessionInfo()
```
